```{r setup, include=FALSE}
knitr::opts_knit$set(verbose = TRUE)
knitr::opts_knit$set(root.dir = "/Users/au698484/Documents/WRIC_processing")
print(getwd())
```

Welcome!

This document serves as a guided tutorial for our workshop on **06.03.2025** at Steno Diabetes Center Aarhus. It is designed to help you get started using the WRIC_preprocessing package, but you can use it afterwards to follow it on your own.

Before we dive into the code-along session, please ensure you have the following installed:


-  **R**: The programming language we’ll be using for data analysis and visualization.
- **A Programming Environment (RStudio)**: I will demonstrate using RStudio, but feel free to use a different IDE.
- **Quarto**: To render this tutorial, if you choose to go through it by yourself.

<details>
<summary>Click here for installation links and instructions</summary>
- **[Install R](https://cran.r-project.org/mirrors.html)**
- **[Install RStudio](https://posit.co/download/rstudio-desktop/)**
- **[Install Quarto](https://quarto.org/docs/get-started/)**
</details>

# FAQ and "Programming Terms"
### What is the difference between R and RStudio?
**R** is a programming language, while **RStudio** is a, so called, integrated development environment (IDE) designed specifically for R, offering a user-friendly interface for coding, plotting, and managing projects. You can think of R like a language, like english, where RStudio is Word - a program you can write english inside. But you could also use other programs for example libre office or latex.

### qmd vs R vs Rmd 
There are different types of files where you can write R code. A file ending in **.R** is a standard R script for writing and running R code, while in **.qmd (Quarto Document)** or **.Rmd (R Markdown)** you can combine code and text for dynamic reports, or for example this tutorial.


# Getting Started
All functions are contained in the `R/WRIC_preprocessing.R` file. But to also have access to example data, doc_strings, tutorials and HowTo's please download the entire repository for this workshop.

1. Download the entire repository from GitHub
- Go to the [repository on GitHub](https://github.com/hulmanlab/WRIC_processing) and click the **Code** button.
- Click **Download ZIP** to download the entire repository, then extract it.
If you have Git installed on your computer or GitHub Desktop, feel free to clone the repository instead of downloading it. 
`git clone https://github.com/hulmanlab/WRIC_processing.git`

2. Set your working directory
- Set your working directory in R to the downloaded WRIC_processing directory. That means your path should end in `".../WRIC_processing"`. You can check the full path, by [add info for both mac and windows]
- **Via RStudio menu:** Session -> Set Working Directory -> Choose Directory
- **Via Terminal:**
```
setwd("path/to/this/folder/WRIC_processing")
getwd() # To check the current directory
```

<details>
<summary>Why do we change the working directory</summary>
The working directory is where RStudio is looking for files. Meaning we can easily type source('R/WRIC_preprocessing.R') instead of the entire file path. There is nothing wrong with specifiying the entire file path and not changing the working directory. But for this workshop it makes it a lot easier, when we can all use the same code and this ensure that all code runs on your computer without having to change anything.
</details>

3. Importing the functions from the "package"
- `source()` runs an R script and loads its functions into your environment

```{r}
source("R/WRIC_processing.R")
```


# Preprocess WRIC data
- code chunk with example data (explain what it all does)
- now use parameteres available
- show docstring to see parameter documentation
- play with various parameters

# Batch Processing
- looping over folder with multiple files
- super quick excursion: RedCap API -> verweis auf seperates tutorial

# Automatic note file extraction - adaptation to your notes
- explain how notefile parameter works with example (e.g. necessary to create example data for that)

# Working with a subset of the data (specific time)
- example from already extracted data, import it again as dataframe
- extract certain time frame
- maybe some examplary biostats example? -> ask chatgpt for something simple

# Visualizing WRIC data


```{r}
2 + 2
```
 